---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spagg

<!-- badges: start -->
<!-- badges: end -->

The goal of spagg is to provide tools for aggregating spatial summary statistics generated from
multiple regions-of-interest (ROIs) imaged using spatial proteomics from the same tissue sample. 

## Installation

You can install the development version of spagg from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("sarahsamorodnitsky/spagg")
```

## Example

```{r example}
library(spagg)
```

Let's consider analyzing a non-small cell lung cancer dataset generated from a multiplexed immunohistochemistry study by Johnson et al. (2021). Johnson et al. found that CD4+ T cells and tumor cells colocalized differently in major histocompatibility complex II (MHCII)-high tumors than in MHCII-low tumors. We are going to test the same hypothesis using bivariate Ripley's K as our measure of spatial colocalization between CD4+ T cells and tumor cells. 

Since this dataset consists of multiple ROIs per tumor sample of $n=153$ tumors, we need to aggregate Ripley's K within a sample. `spagg` provides several ways of doing so. 

First, let's load in the data we will use.

```{r data}
data(lung_df_tumor)
```




