---
title: "spagg"
output: rmarkdown::html_vignette
bibliography: bibliography.bib  
vignette: >
  %\VignetteIndexEntry{spagg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
use-biblatex: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we will demonstrate how to use the `spagg` package to aggregate replicated spatial summary statistics estimated across multiple regions-of-interest (ROIs) imaged from the same biospecimen, such as a tumor sample. 

```{r setup, message = FALSE}
# Loading in packages
library(spagg)
library(dplyr)
library(magrittr)
library(spatstat)
```

Let's consider using `spagg` to analyze a multiplexed immunohistochemistry (mIHC) dataset generated from a non-small cell lung cancer study by @johnson2021cancer. Let's examine the spatial distribution of CD4 T cells and examine its association with major histocompatibility complex II (MHCII) status, a tumor-level label. We will use Ripley's K to characterize the spatial distribution of CD4+ T cells. 

Since this study cpatured multiple ROIs per tumor sample we need a way of handling multiple estimates of Ripley's K for each sample. `spagg` provides several ways of doing so. 

First, let's load in the data we will use. This dataset was originally downloaded from [here] (http://juliawrobel.com/MI_tutorial/MI_Data.html) and cleaned for the purposes of this illustration. 
```{r data}
data(lung_df_tumor)
```

We will first consider different approaches to averaging Ripely's K across ROIs within a tumor sample. Let's start by defining some notation. Suppose for sample $i$ we have $R_i$ ROIs. Let $n_{ir}$ be the number of cells in sample $i$ ROI $r$. Let $A_{ir}$ represent the area of sample $i$ ROI $r$. Let $\hat K_{ir}$ represent the spatial distribution of CD4+ T cells evaluate at radius $30$. 

`spagg` contains implementations for the following three weighted averages:

(1) 
\begin{equation*}
  \bar K_i^{Diggle} = \frac{1}{\sum_{r=1}^{R_i} n_{ir} \sum_{r=1}^{R_i} n_{ir} \hat K_{ir}
\end{equation*}

(2)
\begin{equation*}
  \bar K_i^{Baddeley} = \frac{1}{\sum_{r=1}^{R_i} n^2_{ir} \sum_{r=1}^{R_i} n^2_{ir} \hat K_{ir} \end{equation*}

(3)
\begin{equation*}
  \bar K_i^{Landau} = \frac{\sum_{r=1}^{R_i} A_{ir}}{\sum_{r=1}^{R_i} n_{ir}} \sum_{r=1}^{R_i} \frac{n^2_{ir}}{A_{ir}} \hat K_{ir} 
\end{equation*}

$\bar K_i^{Diggle}$ (@diggle1991analysis), $\bar K_i^{Baddeley}$ (@baddeley1993analysis), and $\bar K_i^{Landau}$ (@landau2004nonparametric) were each proposed as approaches to aggregating replicated Ripley's K statistics.  

Let's estimate Ripley's K for $r=30$ on each ROI.

```{r Kest}
# Set the radius
r <- 30

# Save the image IDs
ids <- unique(lung_df_tumor$id)

# Initialize a data.frame to store the results for each ROI
results <- lung_df_tumor %>% 
  dplyr::select(PID, id) %>%
  dplyr::distinct() %>%
  dplyr::mutate(spatial = NA,
                npoints = NA,
                area = NA)
 
# Iterate through the ROIs (this will take a few seconds)
for (i in 1:length(ids)) {
  
 # Save the ith image
 image.i <- lung_df_tumor %>%
   dplyr::filter(id == ids[i]) %>%
   dplyr::select(x, y, type)

 # Convert to a point process object
 w <- spatstat.geom::convexhull.xy(image.i$x, image.i$y)
 image.ppp <- spatstat.geom::as.ppp(image.i, W = w, marks = image.i$type)
 image.ppp <- subset(image.ppp, marks %in% "CD4+ T cell")

 # Compute Kest
 Ki <- spatstat.explore::Kest(image.ppp, r = 0:r)

 # Calculate the number of points
 npoints.i <- spatstat.geom::npoints(image.ppp)
 
 # Calculate the area
 area.i <- spatstat.geom::area(image.ppp)

 # Save the results
 results[results$id == ids[i],]$spatial <- Ki$iso[r+1]
 results[results$id == ids[i],]$npoints <- npoints.i
 results[results$id == ids[i],]$area <- area.i
}

# View the first few rows
head(results)
```

Some of the Ripley's K estimates will be `NaN`s as a result of there being only $0$ or $1$ CD4 + T cell. We will remove these ROIs from our analysis. After that, we will compute the weighted averages given above. We will also compute a standard arithmetic mean as a comparison. 

```{r averages}
# Remove NAs
results <- results[!is.na(results$spatial),]

# Compute averages
results_mean <- results %>%
    dplyr::group_by(PID) %>%
    dplyr::summarise_at(dplyr::vars(spatial),
                        list(~mean(.x), 
                             ~landau.avg(K.vec = .x, area.vec = area, n.vec = npoints),
                             ~diggle.avg(K.vec = .x, n.vec = npoints),
                             ~baddeley.avg(K.vec = .x, n.vec = npoints)))
```

We now have a data.frame with a column corresponding to each average. Let's add in the MHCII-high status variable so we can do association testing. 

```{r testing with averages}
# First, check the ordering matches
all(unique(lung_df_tumor$PID) == results_mean$PID) # TRUE!

# Add in MHCII-high status
results_mean$mhcII_high <- lung_df_tumor %>% 
  dplyr::select(PID, mhcII_high) %>%
  dplyr::distinct() %>%
  dplyr::select(mhcII_high) %>%
  unlist()

# Test for an association between CD4 T cell spatial distributions and MHCII-high status
mean.glm <- glm(mhcII_high ~ mean, data = results_mean, family = binomial())
diggle.glm <- glm(mhcII_high ~ diggle.avg, data = results_mean, family = binomial())
baddeley.glm <- glm(mhcII_high ~ baddeley.avg, data = results_mean, family = binomial())
landau.glm <- glm(mhcII_high ~ landau.avg, data = results_mean, family = binomial())

# P-values from each approach
summary(mean.glm)$coef[2,4]
summary(diggle.glm)$coef[2,4]
summary(baddeley.glm)$coef[2,4]
summary(landau.glm)$coef[2,4]
```

We did not find any significant associations between the spatial distribution of CD4 T cells as estimated by Ripley's K and MHCII-high status at a radius of $r=30$. 

We now illustrate using ensemble testing to test the same hypothesis. `spagg` contains implementations for three different approaches:


- Standard ensemble testing: in this approach, `spagg` randomly generates weights from a standard normal distribution which are used to construct a weighted mean of the Ripley's K estimates obtained from each ROI. This process is repeated many times. For each random set of weights, the association between the weighted mean spatial summary and sample-level outcomes is tested. The resulting p-values are combined using the Cauchy combination test. This approach is contained in the `ensemble.avg` function. 

- Resampling: in this approach, `spagg` randomly samples one spatial summary statisti per sample and tests for an association with the sample-level outcome. This process is repeated many times and the resulting p-values are combined using the Cauchy combination test. This approach is contained in the `resampling.avg` function. 

- Combination testing: in this approach, `spagg` randomly generates weights from a standard normal distribution and uses these weights as well as the number of cells in each ROI to construct a weighted mean of Ripley's K estimates. This process is repeated many times and the resulting p-values are combined using the Cauchy combination test. This approach is contained in the `combo.weight.avg` function. 

We now run each of these approaches to testing.

```{r ensemble}
# Add the outcome back into the data
results <- left_join(results, 
                     lung_df_tumor %>% dplyr::select(id, PID, mhcII_high) %>% distinct(), 
                     by = c("id", "PID"))

# Run each ensemble test
ensemble.res <- ensemble.avg(data = results, group = "PID", 
                             outcome = "mhcII_high", model = "logistic")

resample.res <- resampling.avg(data = results, group = "PID", 
                               outcome = "mhcII_high", model = "logistic")

combo.res <- combo.weight.avg(data = results, group = "PID", 
                              outcome = "mhcII_high", model = "logistic")

# P-values
ensemble.res$pval
resample.res$pval
combo.res$pval
```

The ensemble p-values also show no significance in the association between the spatial clustering of CD4+ T cells and MHCII-high status among tumor samples. 

# References

<div id="refs"></div>

